---
title: "NADA2"
output: github_document
---

```{r setup, include=FALSE,echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Table of Contents

* [Introduction](#intro)
* [Citing package](#cite)
* [Installation](#install)

## Introduction <a name="intro"></a>
For many environmental professionals, the way to deal with "nondetects" is less than obvious. Values below detection or reporting limits result from measuring trace amounts of a variety of organic and inorganic chemicals. This package includes (maximum likelihood and survival analysis) methods for computing summary statistics, hypothesis tests, and regression for data with one or more detection limits. 

This package is a series of new procedures and improvements on current ones in the `NADA` package developed through a series of workshops and updates to Helsel (2004).

- Helsel, D.R., 2004. Nondetects and Data Analysis: Statistics for Censored Environmental Data, 1st edition. ed. Wiley-Interscience, Hoboken, N.J.


## Citing package <a name="cite"></a>

```{r,eval=F}
citation('NADA2')
```


## Installation <a name="install"></a>

```{r,eval=F}
install.packages("devtools");# if you do not have it installed on your PC
devtools::install_github("SwampThingPaul/NADA2")
```


***
